The Emotional Coherence Controversy: Evidence from Human and Automatic Expression Recognition

### 2019-04-21 ACII_2019_paper/ACII_2019_paper.tex
final version

\hypersetup{
            pdftitle={Coherence between subjective feeling and facial expression: A comparison of human and automatic recognition},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}
\usepackage{wrapfig}
\usepackage[none]{hyphenat}
\floatplacement{figure}{H}

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Coherence between subjective feeling and facial expression: A comparison
of human and automatic recognition}

@@

%% ---- classic IEEETrans one column per institution --------
 %% -- beg if/affiliation.institution-columnar
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Damien Dupré %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Dublin City University\\
Dublin, Ireland
\\damien.dupre@dcu.ie
}
\and
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Anna Tcherkassof %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Grenoble Alpes University\\
Grenoble, France
  %% -- beg for/affiliation.institution.author
\\anna.tcherkassof@univ-grenoble-alpes.fr
 %% -- end for/affiliation.institution.author
}
 %% -- end for/affiliation.institution
 %% -- end if/affiliation.institution-columnar
%% ----------------------------------------------------------

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
While it has been taken for granted in the development of several
automatic facial expression recognition tools, the question of the link
between subjective feelings and facial expressions is still a subject of
debate. On one hand the behaviorist approach conceives emotions as
genetically hardwired and therefore being genuinely displayed through
facial expressions. On the other hand the constructivist approach
conceives emotions as socially constructed, the emotional meaning of a
facial expression being inferred by the observer. In order to evaluate
the link between the subjective feeling of emotions and their
recognition based on facial expression, 232 videos of encoders recruited
to carry out an emotion elicitation task were annotated by 1383 human
observers as well as by an automatic facial expression classifier.
Results show a low accuracy of human observers and of the automatic
classifier to infer the subjective feeling from the facial expressions
displayed by encoders . They also show a weak link between self-reported
emotional states and facial emotional displays. Based on these results,
the hypothesis of genetically hardwired emotion genuinely display is
difficult to support whereas the idea of emotion and facial expression
socially constructed appears to be more likely. Then automatic emotion
recognition tools based on facial expressions should be questioned.
\end{abstract}

@@

With the development of commercial automatic facial expression
recognition tools {[}1{]}, industries and governments are gradually
implementing this technology in order to track humans' emotions in
various scenarios (\emph{e.g.}, marketing, healthcare, automotive to
name a few). This technoloogy rests on the premise that facial
expressions provide a direct access to individuals' subjective feeling.
Even if this premise is central to the modern mainstream approach of
human emotion, recent research in affective science are challenging it.
Once the two competing approaches briefly described as well as the
predictions they respectively entail, an experiment testing these
hypotheses will be presented and its results analyzed in order to
provide empirical evidence to contribute to answer the question.

@@

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} {[}2{]}, facial
expressions are conceived as a genuine displays of individuals' inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory (BET) which states that a set of six emotions are
universally displayed and are genetically hardwired not only in humans
{[}3{]} but also in different animal species {[}4{]}. According to this
view, \emph{``when emotions are aroused by perception of a social event,
a set of central commands produce patterned emotion-specific changes in
multiple systems, including {[}\ldots{}{]} facial expressions.''} {[}5,
p. 49{]}. To cope with critics, several amendments have been made to the
BET, increasing the number of basic emotions from six to seven {[}6{]}
as well as adding the concept of ``display rules'' to explain cultural
differences in the management of facial expressions {[}7{]}.

Even if this theory obtained a popular support, it fails to explain how
individuals can feel emotions without expressing them and how
individuals can express emotions without feeling them, in instances in
which display rules cannot be called upon {[}8{]}, {[}9{]}. These
evidences have led to an alternative conception, notably the social
constructivist approach.

@@

Detractors of the Basic Emotion Theory consider emotion not as
genetically hardwired but as a learnt association between a given
situation and an appropriate response {[}10{]}, {[}11{]}. For the
tenants of the constructivist approach, emotions are ``concepts'' based
on past experiences and which are \emph{``a collection of embodied,
whole brain representations that predict what is about to happen in the
sensory environment, what the best action is to deal with impending
events, and their consequences for allostasis''} {[}12, p. 12{]}.
Following this assumption, faces are best conceived as tools displaying
signals in social interactions {[}13{]}. These signals can convey
individuals' motivations and readiness {[}14{]} or social messages
{[}15{]}. Therefore, facial expressions are thought as behaviors which
meaning is inferred by the observer. Findings support this observer
dependence {[}16{]}, {[}17{]}. They show that to make meaning of another
person's facial behavior, the perceiver relies in particular on her/his
knowledge about emotion categories.

@@

Regarding emotional facial expression (EFE) recognition, the behaviorist
approach assertions and the constructivist approach claims lead to two
divergent predictions. The foremost postulates that, when triggered,
each basic emotion is expressed by a prototypical face (non basic
emotions being blends of the basic ones). Basic emotions are easily
recognized by all human observers and emotional states are accessible by
facial measurement. As the recognition of facial expressions is based on
the identification of specific patterns of facial movements, it implies
that EFE recognition of both human observers and automatic classifiers
should be as accurate. The constructivist approach, in contrast, affirms
that facial expressions do not provide a direct access to individuals'
subjective feeling. Even if the face possibly moves during an emotional
episode, facial muscle movements are not linked in a one-to-one manner
to a specific discrete emotional experience {[}18, p. p418{]}. Instead,
emotions are mentally constructed by the perceiver and mental categories
of emotions are needed to accurately categorize facial movements.
Therefore, since there is no emotional prototypical face, one should
expect human observers' superior ability to accurately recognize EFE as
compared to automatic EFE recognition tools (Figure
\ref{fig:models_img}).

@@

The aim of the current paper is to investigate the link between the
subjective feeling of emotions and its recognition from facial
expressions in order to give credit either to the behaviorist plea or to
the constructivist's one. Contrary to most studies using posed and
static EFE, the present study focuses on natural EFE. Spontaneous and
dynamic facial reactions to emotional elicitations are under
consideration to ensure the generalizability of the results to emotional
behaviors in ordinary life.

@@

To evaluate the link between subjective feeling of emotions and their
recognition from facial expressions, encoders were first recruited to
perform an emotion elicitation task while their facial expression was
video recorded. Then, the videos of the encoders' faces were shown to
human observers and were also analysed by an automatic classifier in
order to identify which emotion was displayed.

@@

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/dynemo_img-1.pdf}
\caption{\label{fig:dynemo_img}Example of a front facing recording
synced with the full view of the participant and the elicitation task.
This picture is taken from a pilot with projects collaborators and all
gave a consent for the publication of their photos and videos.}
\end{figure}

@@

Finally, a debriefing session was performed to ensure that encoders were
not durably affected by the emotion elicitation task. The debriefing was
also used to check that encoders did not guess the real purpose of the
experiment (\emph{e.g.}, being filmed while they were performing an
emotional elicitation task) to guarantee facial expressions'
genuineness. All encoders gave their agreement on their data and video
to be processed for research purpose only.

@@

The overall correlation of recognition and non-recognition between
self-reported emotions and human observers recognition is significant
but low (\(r = .24\), 95\% CI \([.19\), \(.29]\), \(t(1384) = 9.21\),
\(p < .001\)). In order to identify differences according the emotional
labels, encoders' subjective feeling is compared with human observers
recognition in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_hr}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_hr-1.pdf}
\caption{\label{fig:confusionMatrix_sr_hr}Confusion matrix of between
the emotion self-reported as being characteristic of the elicitation
with the emotion recognized by the human observers.}
\end{figure}

Each emotion label used to describe encoders' self-reported subjective
feeling (\emph{i.e.}, the label rated with the highest value) is
compared with the emotion labels which were rated with the highest score
by human observers. Results of the confusion matrix show a moderate
agreement between the emotion felt by the encoder during the elicitation
and the emotion recognized by the human observers (Accuracy
\nolinebreak = \nolinebreak 0.43, 95\% CI {[}0.35,0.52{]}; Kappa = 0.19)
except for \emph{disgust} (100\% of the videos self-reported). These
results are far from those classicaly obtained in the literature for
emotional facial expression recognition which ranges between 60\% and
80\% accuracy. However these results are mostly obtained with static
(\emph{i.e.,} pictures) and posed (\emph{i.e.,} displayed by actors)
facial expressions using only 6 emotional labels in a forced-choice
paradigm.

@@

Sensitivity, specificity, precision and F1 score for each emotion
reveals that \emph{happiness} has the highest coherence ratio whereas
\emph{sadness} has the lowest coherence ratio between true positives and
false positives (Table \ref{table:confusionTable_sr_hr}).

@@

Accuracy metrics by emotional labels indicate a discrepancy in the ratio
of true/false positives. Whereas \emph{happiness} and \emph{disgust}
obtain the highest scores, \emph{anger}, \emph{surprise} and
\emph{sadness} have the lowest recognition ratio. The underlying effect
of expression intensity may explain why \emph{happiness} and
\emph{disgust} are easily recognized. \emph{Anger} and \emph{sadness} as
non-socialy desirable emotion may be have been felt but not expressed.

@@

Similarly to the previous analysis, the overall correlation of
recognition and non-recognition revealed a significant but very low link
between self-reported emotions and automatic \nolinebreak classifier's
recognition (\(r = .12\), 95\% CI \([.07\), \(.17]\),
\(t(1384) = 4.50\), \(p < .001\)). A confusion matrix was used to
compare encoders' subjective feeling with the emotion label recognized
by the automatic classifier (Figure \ref{fig:confusionMatrix_sr_ar}).

@@

Results obtained for the comparison between emotions self-reported and
recognized by the automatic classifier are somewhat similar to the ones
with human observers (Table
\nolinebreak \ref{table:confusionTable_sr_ar}). Overall, a low agreement
between emotion self-reported and emotion recognized by the automatic
classifier (Accuracy \nolinebreak = \nolinebreak 0.3, 95\% CI
{[}0.22,0.38{]}; Kappa \nolinebreak = \nolinebreak 0.07) except for
\emph{happiness} (44.4\% of the video self-reported) is watched.

Surprisingly the automatic classifier incorrectly recognized as
\emph{disgust} an important proportion of videos in which \emph{anger},
\emph{happiness} and \emph{surprise} was the highest self-reported
emotion (respectively 42.9\%, 25.0\% and 32.6\% of the videos
self-reported). In parallel, the automatic classifier recognized as
\emph{happiness} videos in which \emph{fear} and \emph{surprise} was the
highest self-reported emotion (respectively 66.7\% and 32.6\% of the
videos self-reported).

@@

As previously mentioned, human observers appear to be more accurate than
the automatic classifier to recognize individuals' subjective feeling
(human observers Accuracy \nolinebreak = \nolinebreak0.43; automatic
classifier Accuracy = 0.3; \(r = .22\), 95\% CI \([.17\), \(.27]\),
\(t(1384) = 8.31\), \(p < .001\)). However, both make mistakes.

@@

The overall agreement between human observers and the automatic
classifier is in fact very low (Kappa = 0.18). Except for
\emph{happiness} and \emph{disgust} (respectively 64.9\% and 50.0\% of
common labelling), there is no clear common pattern. Moreover, the
automatic classifier has a tendency to label as \emph{disgust} videos
labeled as \emph{sadness} by human observers and as \emph{happiness}
videos labeled as \emph{fear} by human observers

@@

Despite being one on the most investigated question in affective
science, the coherence between emotion felt and facially displayed and
facial expression recognized is a hot topic. To date no clear evidence
has been found to definitely answer it. Yet, with the growing interest
of industries and government to monitor individual's psychological
states, this issue is under high pressure. The present research aims to
provide some empirical data to the question. The faces encoders
displayed when confronted with an emotional eliciting task were
submitted both to human and to automatic recognition. The criterion for
recognition accuracy was the subjective feeling self-reported by the
encoder once the elicitation task carried out. Results first reveal a
low coherence between emotion felt and facial expression displayed.
Secondly, results show a low accuracy for both humans and automatic
classifier in identifying the inner emotional states of these encoders
based on their facial expressions. Thirdly, human observers prove to be
better at recognizing the emotion facially expressed than the automatic
recognition tool is.

Such results support the hypothesis advanced by some authors of a low
emotion--expression coherence {[}23{]}. In many instances, facial
displays are not associated with a concordant emotional state, even any
emotional state at all {[}24{]}, {[}25{]}. More and more evidences are
showing that facial expressions are in reality not expressing emotions
{[}26{]}. As well as other nonverbal behaviors, facial movements are not
only assumed to be determined by emotion but also by various other
causes, such as psychological states (\emph{e.g.},
\nolinebreak motivations or pain), to say nothing of social context and
sociocultural norms {[}7{]}. This multiple determination excludes any
possibility of drawing an inference from facial activity on the
underlying psychological state (emotional or other).

Beyond the present observations showing a weak coherence between
subjective feelings and spontaneous facial expressions, this study
provides some light on the controversy between the behaviorist and the
constructivist approaches. The first of those two lines of thinking
{[}3{]} assumes that expressions of emotion are brief and coherent
patterns of facial muscle movements that co-vary with discrete
subjective experiences. Emotions and their related prototypical facial
behavior are universal because they are considered as innate mechanisms
allowing individuals to respond adaptively to evolutionary significant
events (threats, opportunities\ldots{}) encountered in the environment.
Instead of viewing emotions as natural kinds {[}12{]}, the
constructivist approach supposes that emotions are social constructions.
The specific emotion categories identified on the face of another person
are the result of categorization. In other words, the emotions that are
recognized by the observer are constructed in her/his mind, primarily
based on her/his conceptual knowledge of emotion. Therefore, facial
movements do not express specific emotions. It is the observer that
infers the emotional meaning of the facial expression. As a consequence,
one can predict from the first line of thinking that individuals'
emotional subjective feeling should be correlated to the recognition of
facial expressions from both human observers and automatic classifiers
whereas if emotions are social constructs, as stated by the second line
of thinking, human observer's should be better at perceiving emotions in
face than automatic classifiers.

Present results plead in favor of the latter stance. They show that
human observers are more accurate than the automatic recognition tool to
identify individuals' subjective feeling on the basis of their face.
Moreover, mistakes made by human observers look less arbitrary to the
ones made by the classifier. For instance, even if mix-up between
disgust and anger is sometimes reported in recognition studies,
confusions such as the ones produced by the classifier have never been
noted for human observers. A possible explanation is that human
observers are assessing dynamic expressive sequences. They have access
to the full video which provides a kinetic context whereas the automatic
classifier only assesses the videos frame by frame. Further research is
needed to shed light on this issue.

Some limitations should be stated, notably regarding the use of
self-reports to evaluate encoders' subjective feelings. Accessing the
inner subjective feeling can be biased if not impossible. Moreover the
procedure used for human observation can also be open to dispute.
Instead of asking the human annotators to provide a unique label, a more
subtle approach was chosen to mimic results provided by the automatic
classifier. Whereas this paradigm is longer and more complicated, it can
lead to more robust results in reducing the forced-choice bais {[}27{]}.
However this procedure can also reduce the human observers accuracy. In
this regard, the results of the human observation could have been more
ambiguous because it is not the natural way that people are inferring
meaning from facial expressions. An alternative explanation relies in
reducing the recognition bais involved in the classic recognition
paradigm. Classic forced-choice paradigms obtain artificial high
results, thus by using a more evolved approach observers' accuracy may
have been lowered.

Considering the above, the results provide additional evidence that
individuals' subjective feeling can not be inferred from facial
expressions and in our case invalidate the hypothesis of hardwired
emotions unambiguously displayed on the face. Even if emotions were
hardwired, in everyday life one does not observe prototypical facial
expressions and therefore research should be focused on analysing
non-prototypical facial expressions. Advancements in identifying
``non-basic'' emotion labels {[}28{]} as well as non-prototypical facial
expression have been made in the developpement of automatic facial
expression recognition tools. However this result suggests that
automatic facial expression recognition tools should merely evaluate
facial morphology features such as action units {[}29{]} rather than
inferring supposedly emotional or affective states.

@@

\hypertarget{references}{%
\section*{VI. References}\label{references}}
\addcontentsline{toc}{section}{VI. References}

@@

>>>>

\hypersetup{
            pdftitle={The Emotional Coherence Controversy: Evidence from Human and Automatic Expression Recognition},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}
\usepackage{wrapfig}
\usepackage[none]{hyphenat}
\floatplacement{figure}{H}
\usepackage{flushend}
\usepackage{biblatex}

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{The Emotional Coherence Controversy: Evidence from Human and Automatic
Expression Recognition}

@@

%% ---- classic IEEETrans one column per institution --------
 %% -- beg if/affiliation.institution-columnar
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Damien Dupré %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Business School\\
Dublin City University\\
Dublin, Ireland
\\damien.dupre@dcu.ie
}
\and
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Anna Tcherkassof %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Laboratoire Inter-universitaire de Psychologie\\
Université Grenoble Alpes\\
Grenoble, France
  %% -- beg for/affiliation.institution.author
\\anna.tcherkassof@univ-grenoble-alpes.fr
 %% -- end for/affiliation.institution.author
}
 %% -- end for/affiliation.institution
 %% -- end if/affiliation.institution-columnar
%% ----------------------------------------------------------

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
While it has been taken for granted in the development of several
automatic facial expression recognition tools, the question of the
coherence between subjective feelings and facial expressions is still a
subject of debate. On one hand, the behaviorist approach conceives
emotions as genetically hardwired and therefore being genuinely
displayed through facial expressions. On the other hand, the
constructivist approach conceives emotions as socially constructed; the
emotional meaning of a facial expression being inferred by the observer.
In order to evaluate the coherence between the subjective feeling of
emotions and their recognition based on facial expression, 232 videos of
encoders recruited to carry out an emotion elicitation task were
annotated by 1383 human observers as well as by an automatic facial
expression classifier. Results show low accuracy of human observers and
of the automatic classifier to infer the subjective feeling from the
facial expressions displayed by encoders. They also show a weak
coherence between self-reported emotional states and facial emotional
displays. Based on these results, the hypothesis of genetically
hardwired emotion genuinely displayed is difficult to support, whereas
the idea of emotion and facial expression socially constructed appears
to be more likely. Accordingly, automatic emotion recognition tools
based on facial expressions should be questioned.
\end{abstract}

@@

With the development of commercial automatic facial expression
recognition tools {[}1{]}, industries and governments are gradually
implementing this technology in order to track humans' emotions in
various scenarios (\emph{e.g.}, in marketing, healthcare, and the
automotive industry to name a few). This technology rests on the premise
that facial expressions provide a direct access to individuals'
subjective feelings also called ``emotional coherence''. Even if this
premise is central to the modern mainstream approach of human emotion,
recent research in affective science is challenging it. Once the two
competing approaches are briefly described along with the predictions
they respectively entail, an experiment testing these hypotheses will be
presented and its results analyzed in order to provide empirical
evidence to contribute to answer the question.

@@

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} {[}2{]}, facial
expressions are conceived as a genuine displays of an individual's inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory (BET) which states that a set of six emotions are
universally displayed and are genetically hardwired not only in humans
{[}3{]} but also in different animal species {[}4{]}. According to this
view, \emph{``when emotions are aroused by perception of a social event,
a set of central commands produce patterned emotion-specific changes in
multiple systems, including {[}\ldots{}{]} facial expressions.''} {[}5,
p. 49{]}. To respond to criticisms, several amendments have been made to
the BET, increasing the number of basic emotions from six to seven
{[}6{]} as well as adding the concept of ``display rules'' to explain
cultural differences in the management of facial expressions {[}7{]}.

Even if this theory obtained popular support, it fails to explain how
individuals can feel emotions without expressing them, and how
individuals can express emotions without feeling them, in instances in
which display rules cannot be called upon {[}8{]}, {[}9{]}. These
evidences have led to an alternative conception, notably the social
constructivist approach.

@@

Detractors of the Basic Emotion Theory consider emotion not as
genetically hardwired but as a learnt association between a given
situation and an appropriate response {[}10{]}, {[}11{]}. For the
proponents of the constructivist approach, emotions are ``concepts''
based on past experiences and which are \emph{``a collection of
embodied, whole brain representations that predict what is about to
happen in the sensory environment, what the best action is to deal with
impending events, and their consequences for allostasis''} {[}12, p.
12{]}. Following this assumption, faces are best conceived as tools
displaying signals in social interactions {[}13{]}. These signals can
convey individuals' motivations and readiness {[}14{]} or social
messages {[}15{]}. Therefore, facial expressions are thought as
behaviors which meaning is inferred by the observer. Findings support
this observer dependence {[}16{]}, {[}17{]}. They show that to make
meaning of another person's facial behavior, the perceiver relies in
particular on her/his knowledge about emotion categories.

@@

Regarding emotional facial expression (EFE) recognition, behaviorist and
constructivist approaches lead to two divergent predictions. The former
postulates that, when triggered, each basic emotion is expressed by a
prototypical face (non basic emotions being blends of the basic ones).
Basic emotions are easily recognized by all human observers and
emotional states are accessible by facial measurement. As the
recognition of facial expressions is based on the identification of
specific patterns of facial movements, it implies that EFE recognition
of both human observers and automatic classifiers should be as accurate.
The constructivist approach, in contrast, affirms that facial
expressions do not provide a direct access to individuals' subjective
feelings. Even if the face possibly moves during an emotional episode,
facial muscle movements are not linked in a one-to-one manner to a
specific discrete emotional experience {[}18, p. p418{]}. Instead,
emotions are mentally constructed by the perceiver and mental categories
of emotions are needed to accurately categorize facial movements.
Therefore, since there is no emotional prototypical face, one should
expect human observers' superior ability to accurately recognize EFE as
compared to automatic EFE recognition tools (Figure
\ref{fig:models_img}).

@@

The aim of the current paper is to investigate the coherence between the
subjective feeling of emotions and its recognition from facial
expressions in order to give credit either to the behaviorist plea or to
the constructivist's one. Contrary to most studies using posed and
static EFE, the present study focuses on natural EFE. Spontaneous and
dynamic facial reactions to emotional elicitations are under
consideration to ensure the generalizability of the results to emotional
behaviors in ordinary life.

@@

To evaluate the coherence between subjective feeling of emotions and
their recognition from facial expressions, encoders were first recruited
to perform an emotion elicitation task while their facial expression was
video recorded. Then, the videos of the encoders' faces were shown to
human observers and were also analysed by an automatic classifier in
order to identify which emotion was displayed.

@@

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/dynemo_img-1.pdf}
\caption{\label{fig:dynemo_img}Example of a front facing recording
synced with the full view of the participant and the elicitation task.
This picture is taken from a pilot with projects collaborators and all
gave consent for the publication of their photos and videos.}
\end{figure}

@@

Finally, a debriefing session was performed to ensure that encoders were
not durably affected by the emotion elicitation task. The debriefing was
also used to check that encoders did not guess the real purpose of the
experiment (\emph{e.g.}, being filmed while they were performing an
emotional elicitation task) to guarantee facial expressions'
genuineness. All encoders gave their agreement on their data and video
to be processed for research purposes only.

@@

The overall correlation of recognition and non-recognition between
self-reported emotions and human observers recognition is significant
but low (\(r = .24\), 95\% CI \([.19\), \(.29]\), \(t(1384) = 9.21\),
\(p < .001\)). In order to identify differences according to the
emotional labels, encoders' subjective feelings are compared with human
observers' recognition in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_hr}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_hr-1.pdf}
\caption{\label{fig:confusionMatrix_sr_hr}Confusion matrix between the
emotion self-reported as being characteristic of the elicitation with
the emotion recognized by the human observers.}
\end{figure}

Each emotion label used to describe encoders' self-reported subjective
feeling (\emph{i.e.}, the label rated with the highest value) is
compared with the emotion labels which were rated with the highest score
by human observers. Results of the confusion matrix show a moderate
agreement between the emotion felt by the encoder during the elicitation
and the emotion recognized by the human observers (Accuracy
\nolinebreak = \nolinebreak 0.43, 95\% CI {[}0.35,0.52{]}; Kappa = 0.19)
except for \emph{disgust} (100\% of the videos self-reported). These
results are far from those classically obtained in the literature for
emotional facial expression recognition which ranges between 60\% and
80\% accuracy. However these results are mostly obtained with static
(\emph{i.e.,} pictures) and posed (\emph{i.e.,} displayed by actors)
facial expressions using only 6 emotional labels in a forced-choice
paradigm.

@@

Sensitivity, specificity, precision and F1 scores for each emotion
reveals that \emph{happiness} has the highest coherence ratio whereas
\emph{sadness} has the lowest coherence ratio between true positives and
false positives (Table \ref{table:confusionTable_sr_hr}).

@@

Accuracy metrics by emotional labels indicate a discrepancy in the ratio
of true/false positives. Whereas \emph{happiness} and \emph{disgust}
obtain the highest scores, \emph{anger}, \emph{surprise} and
\emph{sadness} have the lowest recognition ratio. The underlying effect
of expression intensity may explain why \emph{happiness} and
\emph{disgust} are easily recognized. \emph{Anger} and \emph{sadness} as
non-socially desirable emotions may be have been felt but not expressed.

@@

Similarly to the previous analysis, the overall correlation of
recognition and non-recognition revealed a significant but very low
coherence between self-reported emotions and automatic
\nolinebreak classifier's recognition (\(r = .12\), 95\% CI \([.07\),
\(.17]\), \(t(1384) = 4.50\), \(p < .001\)). A confusion matrix was used
to compare encoders' subjective feeling with the emotion label
recognized by the automatic classifier (Figure
\ref{fig:confusionMatrix_sr_ar}).

@@

Results obtained for the comparison between emotions self-reported and
recognized by the automatic classifier are somewhat similar to the ones
with human observers (Table
\nolinebreak \ref{table:confusionTable_sr_ar}). Overall, a low agreement
between emotion self-reported and emotion recognized by the automatic
classifier (Accuracy \nolinebreak = \nolinebreak 0.3, 95\% CI
{[}0.22,0.38{]}; Kappa \nolinebreak = \nolinebreak 0.07) except for
\emph{happiness} (44.4\% of the video self-reported) is evident.

Surprisingly the automatic classifier incorrectly recognized as
\emph{disgust} an significant proportion of videos in which
\emph{anger}, \emph{happiness} and \emph{surprise} was the highest
self-reported emotion (respectively 42.9\%, 25.0\% and 32.6\% of the
videos self-reported). In parallel, the automatic classifier recognized
as \emph{happiness} videos in which \emph{fear} and \emph{surprise} was
the highest self-reported emotion (respectively 66.7\% and 32.6\% of the
videos self-reported).

@@

As previously mentioned, human observers appear to be more accurate than
the automatic classifier to recognize an individual's subjective feeling
(human observers Accuracy \nolinebreak = \nolinebreak0.43; automatic
classifier Accuracy = 0.3; \(r = .22\), 95\% CI \([.17\), \(.27]\),
\(t(1384) = 8.31\), \(p < .001\)). However, both make mistakes.

@@

The overall agreement between human observers and the automatic
classifier is in fact very low (Kappa = 0.18). Except for
\emph{happiness} and \emph{disgust} (respectively 64.9\% and 50.0\% of
common labelling), there is no clear common pattern. Moreover, the
automatic classifier has a tendency to label as \emph{disgust} videos
labeled as \emph{sadness} by human observers, and as \emph{happiness}
videos labeled as \emph{fear} by human observers.

@@

Despite being one of the most investigated questions in affective
science, the coherence between emotion felt and facially displayed on
the one hand, and facial expression recognized on the other is a hot
topic. To date no clear evidence has been found to definitively solve
the questions raised. Yet, with the growing interest of industries and
government in monitoring individual's psychological states, this issue
is under intense scrutiny. The present research aims to provide some
empirical data to answer some of the questions posed. The faces encoders
displayed when confronted with an emotional eliciting task were
submitted both to human and to automatic recognition. The criterion for
recognition accuracy was the subjective feeling self-reported by the
encoder once the elicitation task was carried out. Results first reveal
a low coherence between emotion felt and facial expression displayed.
Secondly, results show low accuracy rates for both humans and the
automatic classifier in identifying the inner emotional states of these
encoders based on their facial expressions. Thirdly, human observers
prove to be better at recognizing the emotion facially expressed than
the automatic recognition tool is.

Such results support the hypothesis advanced by some authors of a low
emotion--expression coherence {[}23{]}. In many instances, facial
displays are not associated with a concordant emotional state, even any
emotional state at all {[}24{]}, {[}25{]}. More and more evidence is
showing that facial expressions are in reality not expressing emotions
{[}26{]}. As well as other nonverbal behaviors, facial movements are not
only assumed to be determined by emotion but also by various other
causes, such as psychological states (\emph{e.g.},
\nolinebreak motivations or pain), to say nothing of social context and
sociocultural norms {[}7{]}. This multiple determination excludes any
possibility of drawing an inference from facial activity on the
underlying psychological state (emotional or other).

Beyond the present observations showing a weak coherence between
subjective feelings and spontaneous facial expressions, this study sheds
some light on the controversy between the behaviorist and the
constructivist approaches. The first of those two lines of thinking
{[}3{]} assumes that expressions of emotion are brief and coherent
patterns of facial muscle movements that co-vary with discrete
subjective experiences. Emotions and their related prototypical facial
behavior are universal because they are considered as innate mechanisms
allowing individuals to respond adaptively to evolutionary significant
events (threats, opportunities\ldots{}) encountered in the environment.
Instead of viewing emotions as natural kinds {[}12{]}, the
constructivist approach supposes that emotions are social constructions.
The specific emotion categories identified on the face of another person
are the result of categorization. In other words, the emotions that are
recognized by the observer are constructed in her/his mind, primarily
based on her/his conceptual knowledge of emotion. Therefore, facial
movements do not express specific emotions. It is the observer that
infers the emotional meaning of the facial expression. As a consequence,
one can predict from the first line of thinking that individuals'
emotional subjective feeling should be correlated to the recognition of
facial expressions from both human observers and automatic classifiers
whereas if emotions are social constructs, as stated by the second line
of thinking, human observer's should be better at perceiving emotions
expressed on the face than automatic classifiers.

Present results plead in favor of the latter stance. They show that
human observers are more accurate than the automatic recognition tool to
identify an individual's subjective feeling on the basis of their face.
Moreover, mistakes made by human observers look less arbitrary to the
ones made by the classifier. For instance, even if a mix-up between
disgust and anger is sometimes reported in recognition studies,
confusions such as the ones produced by the classifier have never been
noted for human observers. A possible explanation is that human
observers are assessing dynamic expressive sequences. They have access
to the full video which provides a kinetic context whereas the automatic
classifier only assesses the videos frame by frame. Further research is
needed to shed light on this issue.

Some limitations should be stated, notably regarding the use of
self-reports to evaluate encoders' subjective feelings. Accessing the
inner subjective feeling can be biased if not impossible. Moreover the
procedure used for human observation can also be open to dispute.
Instead of asking the human annotators to provide a unique label, a more
subtle approach was chosen to mimic results provided by the automatic
classifier. Whereas this paradigm is longer and more complicated, it can
lead to more robust results in reducing the forced-choice bias {[}27{]}.
However this procedure can also reduce the human observers' accuracy. In
this regard, the results of the human observation could have been more
ambiguous because it is not the natural way that people are inferring
meaning from facial expressions. An alternative explanation relies in
reducing the recognition bias involved in the classic recognition
paradigm. Classic forced-choice paradigms obtain artificially high
results, thus by using a more evolved approach observers' accuracy may
have been lowered.

Considering the above, the results provide additional evidence that an
individual's subjective feeling can not be inferred from facial
expressions and in our case invalidate the hypothesis of hardwired
emotions unambiguously displayed on the face. Even if emotions were
hardwired, in everyday life one does not observe prototypical facial
expressions and therefore research should be focused on analysing
non-prototypical facial expressions. Advancements in identifying
``non-basic'' emotion labels {[}28{]} as well as non-prototypical facial
expression have been made in the development of automatic facial
expression recognition tools. However, these results suggest that
automatic facial expression recognition tools should merely evaluate
facial morphology features such as action units (already evaluated in
OpenFace {[}29{]}, Affectiva's Affdex {[}21{]} or Vicar Vision's
FaceReader {[}30{]} to name a few) rather than inferring supposedly
emotional or affective states.

@@

\hypertarget{references}{%
\section{VI. References}\label{references}}

\footnotesize

@@

\leavevmode\hypertarget{ref-den2005facereader}{}%
{[}30{]} M. Den Uyl and H. Van Kuilenburg, ``The facereader: Online
facial expression recognition,'' in \emph{Proceedings of measuring
behavior}, 2005, vol. 30, pp. 589--590.
